---
title: "hw02-gapminder"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exploring the gapminder object:

##### Determine type and class of gapminder:
```{r gapminder_object_type}
library("gapminder")
library("tidyverse")
library("ggplot2")

typeof(gapminder) # gives a more general result
class(gapminder)
```

gapminder object is a list of class data frame (data frame columns can be different types)

##### Determine number of rows and columns:
```{r gapminder_dimensions_part1}
nrow(gapminder) # number of observations
ncol(gapminder) # number of variables
```

1704 rows/observations and 6 columns/variables

##### Other ways to determine these:
```{r gapminder_dimensions_part2}
str(gapminder)
dim(gapminder)
```

`str()` gives more information about the object - is useful when want to explore the object in detail; `dim()` gives info about the number of rows and cols together, while `nrow()/ncol()` is useful when want to only look at the number of either one.

##### Data type of each variable:
```{r variable_types}
str(gapminder)
```

- country is a factor
- continent is a factor
- year is integer
- lifeExp is numeric
- pop is integer
- gdpPercap is numeric

### Exploring individual variables:
##### Categorical variable:

```{r continent_variable_exploration}
levels(gapminder$continent)
summary(gapminder$continent)
ggplot(gapminder, aes(continent, fill = continent)) + # fill parameter allows to colour each bar
  geom_bar()
```

- gapminder has 5 values in continent variable: Africa, Americas, Asia, Europe, Oceania.
- The most common level is Africa (found 624 times), the least encountered is Oceania (24 observations).

##### QUantitative variable:

```{r pop_variable_exploration}
summary(gapminder$pop)
ggplot(gapminder, aes(pop)) +
  geom_histogram(aes(y=..density..), bins=50) +
  geom_density()
```

The minimum population value is 60000, max - 1.3e9. Median value is 7.024e6.

### Exploring various plot types:

##### Violin plot of life expectancy for each continent during 1980-1990:

```{r lifeExp_vs_continent_filteredbyyear}
gapminder %>%
  select(continent, lifeExp, year) %>% 
  filter(year >= 1980, 
         year <= 1990) %>% 
  ggplot(aes(continent, lifeExp, colour = continent)) +
  geom_violin() +
  geom_jitter()
```

Note that Oceania has less data points than other continents.

##### Plot of countries in Oceania - determining the trend in population number over the years:

```{r Oceania_scaled_pop_vs_year}
gapminder %>% 
  select(country, continent, year, pop) %>% 
  filter(continent == "Oceania") %>% 
  ggplot(aes(year, pop, colour = country)) +
  geom_point() +
  scale_y_log10()
```

New Zealand has lower population numbers than Australia; for both countries the population was increasing over the years.

##### Plot of countries in Oceania - gdpPercap over the years:

```{r Oceania_gdpPercap_vs_year}
gapminder %>% 
  filter(continent == "Oceania") %>% 
  select(country, year, gdpPercap) %>% 
  ggplot(aes(year, gdpPercap, fill = country)) +
  geom_bar(position = "dodge", stat = "identity") 
# position = "dodge" places 2 bars side by side; stat = "identity" means bar height = data value
```

Before ~1970's, gdp per capita betweeen Australia and New Zealand were similar; after that Australia's gdp per capita became higher than that on New Zealand.

##### Plot of population in countries in Europe in the years 1900-2000:

```{r Europe_country_vs_pop_filteredbyyear}
gapminder %>% 
  filter(year >= 1900,
         year <= 2000,
         continent == "Europe") %>% 
  select(pop, country) %>% 
  ggplot(aes(country, pop)) +
  geom_boxplot(fill = "lightblue") +
  geom_jitter() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) # to rotate x-axis text
```

Germany had the biggest population in 20th century compared to other European countries.

##### Plot of population in 2002 vs 2007:

```{r scaled_pop_vs_year_filtered}
gapminder %>% 
  filter(year == "2002" | year == "2007") %>% 
  select(continent, pop, year) %>% 
  ggplot(aes(as.factor(year), pop, shape = continent)) +
  geom_jitter() +
  scale_y_log10() +
  xlab("year")
```

